Does stomatal patterning in amphistomatous leaves minimize the CO\(_2\) diffusion path length within leaves?

### 2024-01-21 ms/ms.tex
fixed reference issue in ms
revised ms and response to reviewers

\abstract{
Photosynthesis is co-limited by multiple factors depending on the plant
and its environment. These include biochemical rate limitations,
internal and external water potentials, temperature, irradiance, and
carbon dioxide (CO\(_2\)). Amphistomatous leaves have stomata on both
abaxial and adaxial leaf surfaces. This feature is considered an
adaptation to alleviate CO\(_2\) diffusion limitations in productive
environments where other factors are not limiting as the diffusion path
length from stomate to chloroplast is effectively halved. Plants can
also reduce CO\(_2\) limitations through other aspects of optimal
stomatal anatomy: stomatal density, distribution, patterning, and size.
A number of studies have demonstrated that stomata are overdispersed on
a single leaf surface; however, much less is known about stomatal
anatomy in amphistomatous leaves, especially the coordination between
leaf surfaces, despite their prevelance in nature and near ubiquity
among crop species. Here we use novel spatial statistics based on
simulations and photosynthesis modeling to test hypotheses about how
amphistomatous plants may optimize CO\(_2\) limitations in the model
angiosperm \emph{Arabidopsis thaliana} grown in different light
environments. We find that 1) stomata are overdispersed, but not ideally
dispersed, on both leaf surfaces across all light treatments; 2) abaxial
and adaxial leaf surface patterning are independent; and 3) the
theoretical improvements to photosynthesis from abaxial-adaxial stomatal
coordination are miniscule (\(\ll 1\)\%) across the range of feasible
parameter space. However, we also find that 4) stomatal size is
correlated with the mesophyll volume that it supplies with CO\(_2\),
suggesting that plants may optimize CO\(_2\) diffusion limitations
through alternative pathways other than ideal, uniform stomatal spacing.
We discuss the developmental, physical, and evolutionary constraits
which may prohibit plants from reaching the theoretical adaptive peak of
uniform stomatal spacing and inter-surface stomatal coordination. These
findings contribute to our understanding of variation in the anatomy of
amphistomatous leaves.}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
movement regulate gas exchange during photosynthesis, namely CO\(_2\)
assimilation and water loss through transpiration. Since waxy cuticles
are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
entry and exit points through which gas exchange occurs despite making
up a small percentage of the leaf area \citep{lange_responses_1971}.
Stomata consist of two guard cells which open and close upon changes in
turgor pressure or hormonal cues \citep{mcadam_linking_2016}. The
stomatal pore leads to an internal space known as the substomatal cavity
where gases contact the mesophyll. Once in the mesophyll, CO\(_2\)
diffuses throughout a network of intercellular air space (IAS) and into
mesophyll cells where CO\(_2\) assimilation (\(A\)) occurs within the
chloroplasts \citep{lee_diffusion_1964}. Stomatal conductance and
transpiration are determined by numerous environmental and anatomical
parameters such as vapor pressure deficit (VPD), irradiance,
temperature, wind speed, leaf water potential, IAS geometry, mesophyll
cell anatomy, and stomatal anatomy.

Many successful predictions about stomata and other C3 leaf traits can
be made by hypothesizing that natural selection should optimize CO\(_2\)
gain per unit of water loss for any given set of environmental
parameters, including their variability
\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
Total stomatal area (size \(\times\) density) is optimized for
operational conductance (\(g_\text{s,op}\)) rather than maximum
conductance (\(g_\text{s,max}\)) such that stomatal apertures are most
responsive to changes in the environment at their operational aperture
\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
can compensate for maladaptive stomatal densities to an extent
\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
determine a leaf's theoretical \(g_\text{s,max}\)
\citep{sack_developmental_2016}, which is proportional to
\(g_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
stomatal densities lead to irregular and insufficient CO\(_2\) supply
and reduced photosynthetic efficiency in leaf areas far from stomata
\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
stomatal densities can reduce water use efficiency (WUE)
\citep{bussis_stomatal_2006} and incur excessive metabolic costs
\citep{deans_optimization_2020}. Stomatal density positively co-varies
with irradiance during leaf development and negatively co-varies with
CO\(_2\) concentration
\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
consistent with optimality predictions. In most species, stomata occur
only on the abaxial (usually lower) leaf surface; but amphistomy, the
occurrence of stomata on both abaxial and adaxial leaf surfaces, is also
prevalent in high light environments with constant or intermittent
access to sufficient water
\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance
\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
Despite these optimality predictions, stomatal anatomy may be partially
constrained by physical and developmental limits on phenotypic
expression
\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.

A number of physical and developmental processes constrain stomatal
anatomy traitspace. For example, almost all stomata follow the one cell
spacing rule to maintain proper stomatal functioning as guard cell
movement requires the rapid exchange of ions with neighboring epidermal
cells (i.e.~subsidiary cells)
\citep{geisler_oriented_2000, dow_physiological_2014}. This would
prevent stomata from being strongly clustered; however, some species
(notably in \emph{Begonia}) appear to benefit from the overlapping vapor
shells caused by stomatal clustering in dry environments
\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
Historically, stomatal patterning in dicot angiosperms was thought to be
random with an exclusionary distance surrounding each stomate
\citep{sachs_developmental_1974}; however, the developmental controls of
stomatal patterning are more complex than random development along the
leaf surface. \citet{croxdale_stomatal_2000} reviews three developmental
theories which attempt to explain stomatal patterning in angiosperms:
inhibition, cell lineage, and cell cycle, ultimately arguing for a cell
cycle based control of stomatal patterning.
\citet{pillitteri_mechanisms_2012} review the short and long distance
signalling pathways associated with stomatal spacing and development,
which include cell to cell communication and whole plant integration to
ensure the proper spacing of stomata across a single leaf surface
depending on environmental ques. Much less is known about the
development of stomata on the adaxial leaf surface in amphistomatous
plants. Stomatal size is additionally constrained by genome size with
larger genomes leading to larger minimum guard cell size
\citep{jordan_environmental_2015}. Despite these limitations,
ecophysiological theory still predicts optimal stomatal anatomy, the
details of which are discussed below.

The patterning and spacing of stomata on the leaf affects photosynthesis
in \(C_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in \(C_3\) plants is generally
co-limited by biochemistry and diffusion, but modulated by light
availability
\citep{parkhurst_intercellular_1990, manter_ci_2004, carriqui_diffusional_2015}.
Low light decreases CO\(_2\) demand by limiting electron transport rate,
leading to relatively high internal CO\(_2\) concentration
(\(C_\text{i}\)) and low \(A_\text{max}\) \citep{kaiser_metabolic_2016}.
In contrast, well hydrated leaves with open stomata in high light,
photosynthesis is often limited by CO\(_2\) supply as resistances from
the boundary layer, stomatal pore, and mesophyll can result in
insufficient CO\(C_2\) supply at the chloroplast to maxmimize
photosynthesis \citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In
this study, we focus primarily on how stomatal patterning affects
diffusion, ignoring boundary layer and mesophyll resistances.

To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
stomata should be uniformly distributed in an equilateral triangular
grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
diffusion path length \citep{parkhurst_diffusion_1994}. As the diffusion
rate of CO\(_2\) though liquid is approximately \(10^4\times\) slower
than CO\(_2\) diffusion through air, mesophyll resistance is generally
thought to be primarily limited by liquid diffusion
\citep{aalto_three-dimensional_2002, evans_resistances_2009}, but
diffusion through the IAS has also been shown to be a rate limiting
process because the tortuous, disjunct nature of the IAS can greatly
increase diffusion path lengths \citep{harwood_understanding_2021}.
Additionally, tortuosity is higher in horizontal directions (parallel to
leaf surface) than vertical directions (perpendicular to leaf surface)
because of the cylindrical shape and vertical arrangement of pallisade
mesophyll cells \citep{earles_beyond_2018, harwood_understanding_2021}.
However, the ratio of lateral to vertical diffusion rate is still
largely unknown
\citep{morison_lateral_2005, pieruschka_lateral_2005, pieruschka_lateral_2006}.
Depending on the thickness of the leaf, porosity of the leaf mesophyll,
tortuosity of the IAS, and lateral to vertical diffusion rate ratio,
minimizing diffusion path length for CO\(_2\) via optimally distributed
stomata may yield significant increases in CO\(_2\) supply for
photosynthesis and higher \(A_\text{max}\).

We hypothesized that natural selection will favor stomatal patterning
and distribution to minimize the diffusion path length. In
amphistomatous leaves, this would be accomplished by 1) a dispersed,
equilateral triangular distribution of stomata on both abaxial and
adaxial leaf surfaces and 2) coordinated stomatal spacing on each
surface that offsets the position of stomata
(\autoref{fig:ideal-amphi-grid}). Coordination between leaf surfaces is
defined in this study as the occurrence of stomata in areas farthest
from stomata on the opposite leaf surface. Additionally, because
CO\(_2\) is more limiting for photosynthesis under high light, we
hypothesize that in high light 3) there should be more stomata, and 4)
stomata should be more uniformly distributed than in low light. Finally,
as stomatal densities are selected for optimal operational aperture, we
hypothesize that 5) stomatal length will be positively correlated with
the area of the leaf surface to which it is closest. We refer to this as
the `stomatal zone', the leaf area surrounding a focal stomate closest
to that stomate and therefore the zone it supplies with CO\(_2\)). This
way, each stomate can be optimally sized relative to the mesophyll
volume it supplies.

@@

To test these hypotheses, we grew the model plant \emph{Arabidopsis
thaliana} in high, medium, and low light and measured stomatal density,
size, and patterning on both leaf surfaces, and spatial coordination
between them. We use Voronoi tessellation techniques to calculate
stomatal zones. We also used a 2-D porous medium approximation of
CO\(_2\) diffusion and photosynthesis to predict the photosynthetic
advantage of optimal versus suboptimal coordination in stomatal
coordination between surfaces. Specifically, we predicted that traits
which affect diffusion path length (leaf thickness, stomatal density,
leaf porosity, lateral-vertical diffusion rate ratio), diffusion rate
(temperature, pressure), and CO\(_2\) demand (Rubisco concentration,
light) would modulate the advantage of optimal stomatal arrangement
following the relationships outlined in \autoref{tab:hypotheses}. Here,
we integrate over reasonable parameter space to determine the
ecophysiological context most likely to favor stomatal spatial
coordination in amphistomatous leaves.

@@

\hypertarget{materials-and-methods}{%
\section{Materials and methods}\label{materials-and-methods}}

\hypertarget{data-preparation}{%
\subsection{Data Preparation}\label{data-preparation}}

Plant material, growth conditions, and three-dimensional confocal
imaging are described in \citet{dow_disruption_2017}. Briefly, Columbia
(Col-0) ecotype of \emph{Arabidopsis thaliana} (L.) Heynh. plants were
grown in three different light environments: low light (PAR = 50
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), and high light (200
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
photosynthetically active radiation. Seeds were surface-sterilized and
stratified at 4°C for 3--5 d in 0.15\% agarose solution and then sown
directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA,
USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer
(Scotts-Sierra, Marysville, OH, USA). At 10--14 d, seedlings were
thinned so only one seedling per container remained. Plants were grown
to maturity in growth chambers where the conditions were as follows: 16
: 8 h, 22 : 20°C, day : night cycle. Imaging of the epidermis and
internal leaf structures was performed using a Leica SP5 confocal
microscope with the protocol developed by
\citet{wuyts_high-contrast_2010} with additional modification described
in \citet{dow_disruption_2017}. We captured 132 images in total, making
66 abaxial-adaxial image pairs. We measured stomatal position and size
using ImageJ \citep{schneider_nih_2012}.

\hypertarget{single-surface-analyses}{%
\subsection{Single surface analyses}\label{single-surface-analyses}}

We tested whether stomata are non-randomly distributed by comparing the
observed stomatal patterning to a random uniform pattern. For each leaf
surface image with \(n\) stomata we generated \(10^3\) synthetic
surfaces with \(n\) stomata uniformly randomly distributed on the
surface. For each sample image, we compared the observed Nearest
Neighbor Index (\(\mathrm{NNI}\)) to the null distribution of
\(\mathrm{NNI}\) values calculated from the synthetic data set.
\(\mathrm{NNI}\) is the ratio of observed mean distance
(\(\overline{D}_O\)) to the expected mean distance (\(\overline{D}_E\))
where \(\overline{D}_E\) is:

@@

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.1 \citep{evans_spatialeco_2023}.
The observed stomatal distribution is dispersed relative to a uniform
random distribution if the observed \(NNI\) is greater than 95\% of the
synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg \citep{benjamini_controlling_1995} false discovery
rate procedure implemented in the \emph{R} package \textbf{multtest}
version 2.56.0 \citep{wong_multiple_2005}.

For each sample image, we also simulated \(10^3\) synthetic data with
\(n\) stomata ideally dispersed in an equilateral triangular grid. For
these grids, we integrated over plausible stomatal densities and then
conditioned on stomatal grids with exactly \(n\) stomata. The simulated
stomatal count was drawn from a Poisson distribution with the mean
parameter \(\lambda\) drawn from a Gamma distribution with shape \(n\)
and scale 1 (\(\lambda \sim \Gamma(n, 1)\)). \(\Gamma(n, 1)\) is the
posterior distribution of \(\lambda\) with a flat prior distribution.
This allows us to integrate over uncertainty in the stomatal density
from the sample image.

We developed a dispersion index \(\mathrm{DI}\) to quantify how close
observed stomatal distributions are to random uniform versus maximally
dispered in an equilateral triangular grid. \(\mathrm{DI}\) varies from
zero to one, where zero is uniformly random and one is ideally
dispersed:

\begin{equation}\label{eq:disp}
  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{uniform}}) - \text{median}(\mathrm{NNI_{random}})}
\end{equation}

@@

\hypertarget{results}{%
\section{Results}\label{results}}

@@

\hypertarget{stomatal-distribution-is-non-random-but-far-from-ideal}{%
\subsection{Stomatal distribution is non-random, but far from
ideal}\label{stomatal-distribution-is-non-random-but-far-from-ideal}}

>>>>

\abstract{
Photosynthesis is co-limited by multiple factors depending on the plant
and its environment. These include biochemical rate limitations,
internal and external water potentials, temperature, irradiance, and
carbon dioxide (CO\(_2\)). Amphistomatous leaves have stomata on both
abaxial and adaxial leaf surfaces. This feature is considered an
adaptation to alleviate CO\(_2\) diffusion limitations in productive
environments as the diffusion path length from stomate to chloroplast is
effectively halved in amphistomatous leaves. Plants may also reduce
CO\(_2\) limitations through other aspects of optimal stomatal anatomy:
stomatal density, distribution, patterning, and size. A number of
studies have demonstrated that stomata are overdispersed compared to a
random distribution on a single leaf surface; however, despite their
prevelance in nature and near ubiquity among crop species, much less is
known about stomatal anatomy in amphistomatous leaves, especially the
coordination between leaf surfaces. Here we use novel spatial statistics
based on simulations and photosynthesis modeling to test hypotheses
about how amphistomatous plants may optimize CO\(_2\) diffusion in the
model angiosperm \emph{Arabidopsis thaliana} grown in different light
environments. We find that 1) stomata are overdispersed, but not ideally
dispersed, on both leaf surfaces across all light treatments; 2) the
patterning of stomata on abaxial and adaxial leaf surfaces is
independent; and 3) the theoretical improvements to photosynthesis from
abaxial-adaxial stomatal coordination are miniscule (\(\ll 1\)\%) across
the range of feasible parameter space. However, we also find that 4)
stomatal size is correlated with the mesophyll volume that it supplies
with CO\(_2\), suggesting that plants may optimize CO\(_2\) diffusion
limitations through alternative pathways other than ideal, uniform
stomatal spacing. We discuss the developmental, physical, and
evolutionary constraits which may prohibit plants from reaching this
theoretical adaptive peak of uniform stomatal spacing and inter-surface
stomatal coordination. These findings contribute to our understanding of
variation in the anatomy of amphistomatous leaves.}

@@

\section{Introduction}\label{introduction}

Stomatal anatomy (e.g.~size, density, distribution, and patterning) and
movement regulate gas exchange during photosynthesis, namely CO\(_2\)
assimilation and water loss through transpiration. Since waxy cuticles
are mostly impermeable to CO\(_2\) and H\(_2\)O, stomata are the primary
entry and exit points through which gas exchange occurs despite making
up a small percentage of the leaf area \citep{lange_responses_1971}.
Stomata consist of two guard cells which open and close upon changes in
turgor pressure or hormonal cues \citep{mcadam_linking_2016}. The
stomatal pore leads to an internal space known as the substomatal cavity
where gases contact the mesophyll. Once in the mesophyll, CO\(_2\)
diffuses throughout a network of intercellular air space (IAS) and into
mesophyll cells where CO\(_2\) assimilation (\(A\)) occurs within the
chloroplasts \citep{lee_diffusion_1964}. Stomatal conductance and
transpiration are determined by numerous environmental and anatomical
parameters such as vapor pressure deficit (VPD), irradiance,
temperature, wind speed, leaf water potential, IAS geometry, mesophyll
cell anatomy, and stomatal anatomy. The latter of these is the focus of
this study, with discussion of other interacting variables.

Many successful predictions about stomata and other C\(_3\) leaf traits
can be made by hypothesizing that natural selection should optimize
CO\(_2\) gain per unit of water loss for any given set of environmental
parameters, including their variability
\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
Total stomatal area (size \(\times\) density) is optimized for
operational conductance (\(g_\text{s,op}\)) rather than maximum
conductance (\(g_\text{s,max}\)) such that stomatal apertures are most
responsive to changes in the environment at their operational aperture
\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
can compensate for suboptimal stomatal densities to an extent
\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
determine a leaf's theoretical \(g_\text{s,max}\)
\citep{sack_developmental_2016}, which is proportional to
\(g_\text{s,op}\) under typical conditions
\citep{mcelwain_using_2016, murray_consistent_2020}. Additionally, low
stomatal densities lead to irregular and insufficient CO\(_2\) supply
and reduced photosynthetic efficiency in leaf areas far from stomata
\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
stomatal densities can reduce water use efficiency (WUE)
\citep{bussis_stomatal_2006} and incur excessive metabolic costs
\citep{de_boer_optimal_2016, deans_optimization_2020}. Stomatal density
positively co-varies with irradiance during leaf development and
negatively co-varies with CO\(_2\) concentration
\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
consistent with optimality predictions. In most species, stomata occur
only on the abaxial (usually lower) leaf surface; but amphistomy, the
occurrence of stomata on both abaxial and adaxial leaf surfaces, is also
prevalent in high light environments with constant or intermittent
access to sufficient water
\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance \citep[
\citet{harrison_influence_2020}]{parkhurst_adaptive_1978, mott_amphistomy_1991}.
Ab- and adaxial leaf surfaces were found to function independent of one
another in wheat, an important crop, with the adaxial surface
demonstrating higher photosynthetic capacity \citep{wall_stomata_2022}.
These results highlight the utmost importance of amphistomy for some
plants.

Despite the success of optimality predictions, stomatal anatomy may be
partially constrained by physical and developmental limits on phenotypic
expression
\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
A number of physical and developmental processes constrain stomatal
anatomy trait space. For example, almost all stomata follow the one cell
spacing rule to maintain proper stomatal functioning as guard cell
movement requires the rapid exchange of ions with neighboring epidermal
cells (i.e.~subsidiary cells)
\citep{geisler_oriented_2000, dow_physiological_2014}. This would
prevent stomata from being strongly clustered; however, some species
(notably in \emph{Begonia}) appear to benefit from the overlapping vapor
shells caused by stomatal clustering in dry environments
\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
Historically, stomatal patterning in eudicot angiosperms was thought to
be random with an exclusionary distance surrounding each stomate
\citep{sachs_developmental_1974}; however, the developmental controls of
stomatal patterning are more complex. \citet{croxdale_stomatal_2000}
reviews three developmental theories which attempt to explain stomatal
patterning in angiosperms: inhibition, cell lineage, and cell cycle,
ultimately arguing for a cell cycle based control of stomatal
patterning. \citet{pillitteri_mechanisms_2012} review the short and long
distance signalling pathways associated with stomatal spacing and
development, which include cell to cell communication and whole plant
integration to ensure the proper spacing of stomata across a single leaf
surface depending on environmental ques. Much less is known about the
development of stomata on the adaxial leaf surface in amphistomatous
plants. Stomatal size is additionally constrained by genome size with
larger genomes leading to larger minimum guard cell size
\citep{jordan_environmental_2015, roddy_scaling_2020}. Despite these
limitations, ecophysiological theory still predicts optimal stomatal
anatomy, the details of which are discussed below.

The patterning and spacing of stomata on the leaf affects photosynthesis
in C\(_3\) leaves by altering the CO\(_2\) diffusion path length from
stomata to sites of carboxylation in the mesophyll. Maximum
photosynthetic rate (\(A_\text{max}\)) in C\(_3\) plants is generally
co-limited by biochemistry and diffusion, but modulated by light
availability
\citep{parkhurst_intercellular_1990, manter_ci_2004, carriqui_diffusional_2015}.
Low light decreases CO\(_2\) demand by limiting electron transport rate,
leading to relatively high internal CO\(_2\) concentration
(\(C_\text{i}\)) and low \(A_\text{max}\) \citep{kaiser_metabolic_2016}.
In contrast, well hydrated leaves with open stomata in high light,
photosynthesis is often limited by CO\(_2\) supply as resistances from
the boundary layer, stomatal pore, sub-stomatal cavity, and mesophyll
can result in insufficient CO\(C_2\) supply at the chloroplast to
maximize photosynthesis
\citep{farquhar_biochemical_1980, lehmeier_cell_2017}. In this study, we
focus primarily on how stomatal patterning affects diffusion.

Assuming uniform mesophyll diffusion resistance in all directions
(homogenous porous medium), an ideal stomatal anatomy can be predicted.
To maximize CO\(_2\) supply from the stomatal pore to chloroplasts,
stomata should be uniformly distributed in an equilateral triangular
grid on the leaf surface so as to minimize stomatal number and CO\(_2\)
diffusion path length \citep{parkhurst_diffusion_1994}. An equilateral
triangular grid is ideal because it maximizes the average distance
between stomata, for a given stomatal density, and thereby minimizes the
average distance between any point in the mesophyll to its nearest
stomate. Assuming a homogenous mesophyll, this is the most efficient
pattern to supply CO\(_2\) to a leaf volume.

Such an assumption, though an oversimplification, is a powerful tool for
photosynthesis modelling, and may provide insight into how real leaves
diverge from this. In real leaves, as the diffusion rate of CO\(_2\)
though liquid is approximately \(10^4\times\) slower than CO\(_2\)
diffusion through air, mesophyll resistance is generally thought to be
primarily limited by liquid diffusion
\citep{aalto_three-dimensional_2002, evans_resistances_2009}, but
diffusion through the IAS has also been shown to be a rate limiting
process because the tortuous, disjunct nature of the IAS can greatly
increase diffusion path lengths \citep{harwood_understanding_2021}.
Additionally, tortuosity is higher in horizontal directions (parallel to
leaf surface) than vertical directions (perpendicular to leaf surface)
because of the cylindrical shape and vertical arrangement of palisade
mesophyll cells \citep{earles_beyond_2018, harwood_understanding_2021}.
However, the ratio of lateral to vertical diffusion rate is still
largely unknown and may be a highly variable trait in leaves
\citep{morison_lateral_2005, pieruschka_lateral_2005, pieruschka_lateral_2006, morison_does_2007}.
Depending on the thickness of the leaf, porosity of the leaf mesophyll,
tortuosity of the IAS, and lateral to vertical diffusion rate ratio,
minimizing diffusion path length for CO\(_2\) via optimally distributed
stomata may yield significant increases in CO\(_2\) supply for
photosynthesis and higher \(A_\text{max}\). Or plants may simply
coordinate the development of stomata and mesophyll IAS to reach another
optimal solution which does not rely on uniformly distributed stomata
\citep{baillie_developmental_2020}.

We hypothesized that, in the absence of any constraint and assuming
homogenous mesophyll diffusion resistance, natural selection will favor
stomatal patterning and distribution to minimize the diffusion path
length. In amphistomatous leaves, this would be accomplished by 1) a
uniform, equilateral triangular distribution of stomata on both abaxial
and adaxial leaf surfaces and 2) coordinated stomatal spacing on each
surface that offsets the position of stomata
(\autoref{fig:ideal-amphi-grid}). Coordination between leaf surfaces is
defined in this study as the occurrence of stomata in areas farthest
from stomata on the opposite leaf surface. Additionally, because
CO\(_2\) is more limiting for photosynthesis under high light, we
hypothesize that in high light 3) there should be more stomata, and 4)
stomata should be more overdispersed (closer to equilateral triangular
grid) compared to a random distribution than in low light. Finally,
since, in measures of whole leaves, stomatal area rather than stomatal
density is optimized for operational conductance, we hypothesize that 5)
stomatal length (and hence its area) will be positively correlated with
the area of the leaf surface to which it is spatial closest as defined
by Voronoi tessellation techniques. We refer to this as the `stomatal
zone', the leaf area surrounding a focal stomate closest to that stomate
and therefore the zone it supplies with CO\(_2\)). This way, each
stomate may be optimally sized relative to the mesophyll volume it
supplies. Hypothesis 3 is already well supported in many species
\citep{poorter_metaanalysis_2019}, but it is useful here to confirm that
light treatments induced plasticity in the expected direction.

@@

To test these hypotheses, we grew the model plant \emph{Arabidopsis
thaliana} in high, medium, and low light and measured stomatal density,
size, and patterning on both leaf surfaces, and spatial coordination
between them. We use Voronoi tessellation techniques to calculate
stomatal zones. We also used a 2-D porous medium approximation of
CO\(_2\) diffusion and photosynthesis to predict the photosynthetic
advantage of optimal versus suboptimal coordination in stomatal
coordination between surfaces. Specifically, we predicted that traits
which affect diffusion path length (leaf thickness, stomatal density,
leaf porosity), diffusion rate (determined by temperature, pressure),
and CO\(_2\) demand (Rubisco concentration, light) would modulate the
advantage of optimal stomatal arrangement following the relationships
outlined in \autoref{tab:hypotheses}. Here, we integrate over reasonable
parameter space to determine the ecophysiological context most likely to
favor stomatal coordination in amphistomatous leaves.

@@

\section{Materials and methods}\label{materials-and-methods}

\subsection{Data Preparation}\label{data-preparation}

Plant material, growth conditions, and three-dimensional confocal
imaging are described in \citet{dow_disruption_2017}. Briefly, Columbia
(Col-0) ecotype of \emph{Arabidopsis thaliana} (L.) Heynh. plants were
grown in three different light environments: low light (PAR = 50
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), medium light (100
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)), and high light (200
\(\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)). PAR stands for
photosynthetically active radiation. \emph{A. thaliana} responds
strongly to light levels over this range
\citep{bailey_acclimation_2001}, though natural populations in open
canopies can experience
\(\text{PAR} > 800~\mu \text{mol}~\text{m}^{-2}~\text{s}^{-1}\)
\citep{callahan_shade-induced_2002}. Seeds were surface-sterilized and
stratified at 4 °C for 3--5 d in 0.15\% agarose solution and then sown
directly into Pro-Mix HP soil (Premier Horticulture; Quakerstown, PA,
USA) and supplemented with Scott's Osmocote Classic 14-14-14 fertilizer
(Scotts-Sierra, Marysville, OH, USA). At 10--14 d, seedlings were
thinned so only one seedling per container remained. Plants were grown
to maturity in growth chambers where the conditions were as follows: 16
: 8 h, 22 : 20°C, day : night cycle. Imaging of the epidermis and
internal leaf structures was performed using a Leica SP5 confocal
microscope (Leica Microsystems, Wetzlar, Germany) with the protocol
developed by \citet{wuyts_high-contrast_2010} with additional
modification described in \citet{dow_disruption_2017}. We captured 132
images in total, making 66 abaxial-adaxial image pairs. Images were
square with an area of 0.386 mm\(^2\). We measured stomatal position and
length using ImageJ \citep{schneider_nih_2012}. A number of synthetic
leaf surface data sets were also simulated (details below) to generate
null distributions against which to test our hypotheses and to avoid any
methodological influence on our results (e.g.~boundary effects when
calculating stomatal patterning). All synthetic leaf surfaces were
simulated based on the size of the real leaf images and stomatal
densities matched those of real leaf images.

\subsection{Single surface analyses}\label{single-surface-analyses}

We compared observed stomatal patterning to an ideal pattern (uniform
equilateral triangular grid) and a null model (random uniform
distribution). The terminology is unfortunately confusing because the
word `uniform' is used in different ways. A uniform equilateral
triangular grid means that the distance between stomata is uniform; a
random uniform distribution means that a stomate has an equal
probability (i.e.~uniform) of occurring anyone on the leaf surface. To
limit confusion, we refer to the ideal pattern as uniform (equilateral
triangle grid) and the null pattern as random (random uniform). When
observed stomatal patterns are more dispersed than expected under random
patterning, we refer to this as overdispersed. Note, however, that
overdispersed compared to random is still less dispersed than ideal,
because the ideal pattern is maximally dispersed.

We tested whether stomata overdispersed by comparing each observed, real
leaf stomatal pattern to an array of synthetic data simulated from a
random distribution. For each observed leaf surface image with \(n\)
stomata we generated \(10^3\) synthetic surfaces with \(n\) stomata
uniformly randomly distributed on the surface. For each sample image, we
compared the observed Nearest Neighbor Index (\(\mathrm{NNI}\)) to the
null distribution of \(\mathrm{NNI}\) values calculated from the
synthetic data set. \(\mathrm{NNI}\) is the ratio of observed mean
distance (\(\overline{D}_O\)) to the expected mean distance
(\(\overline{D}_E\)) where \(\overline{D}_E\) is:

@@

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.2 \citep{evans_spatialeco_2023}.
The observed stomatal distribution is overdispersed relative to a random
distribution if the observed \(NNI\) is greater than 95\% of the
synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg \citep{benjamini_controlling_1995} false discovery
rate procedure implemented in the \emph{R} package \textbf{multtest}
version 2.56.0 \citep{wong_multiple_2005}.

For each sample image, we also simulated \(10^3\) synthetic leaf
surfaces with \(n\) stomata ideally, uniformly dispersed in an
equilateral triangular grid. To account for uncertainty in the stomatal
density of each sample image with \(n\) stomata, we integrated over
plausible stomatal densities and then conditioned on synthetic leaf
surfaces with exactly \(n\) stomata. The simulated stomatal count was
drawn from a Poisson distribution with the mean parameter \(\lambda\)
drawn from a Gamma distribution with shape \(n\) and scale 1
(\(\lambda \sim \Gamma(n, 1)\)). \(\Gamma(n, 1)\) is the posterior
distribution of \(\lambda\) with a flat prior distribution. This
integration was necessary to remove any artifacts of uncertainty in the
true stomatal density of the sample leaves.

We developed a dispersion index \(\mathrm{DI}\) to quantify how close
observed stomatal patterning is to random versus ideally patterned in an
equilateral triangular grid. \(\mathrm{DI}\) varies from zero to one,
where zero is random and one is ideally patterned:

\begin{equation}\label{eq:disp}
  \mathrm{DI} = \frac{\mathrm{NNI} - \text{median}(\mathrm{NNI_{random}})}{\text{median}(\mathrm{NNI_{ideal}}) - \text{median}(\mathrm{NNI_{random}})}
\end{equation}

@@

\section{Results}\label{results}

@@

\subsection{Stomatal patterning is non-random, but far from
uniform}\label{stomatal-patterning-is-non-random-but-far-from-uniform}

Many leaf surfaces (37 of 132, 28\%) are significantly overdispersed
compared to a random uniform distribution, but none were close to an
ideal, uniform equilateral triangular pattern (dispersion index = 1;
\autoref{fig:single-surface}). Before controlling for multiple
comparisons, 43.2\% are significantly overdispersed. The dispersion
index differs significantly among light treatments (ANOVA,
\(F_{2,126} = 8.55, P = 3.30 \times 10^{-4}\)) because the medium light
treatment is significantly less than the low treatment
(\autoref{fig:single-surface}). Dispersion index is consistently greater
on adaxial leaf surfaces across all light treatments (ANOVA,
\(F_{1,126} = 28.8, P = 3.67 \times 10^{-7}\);
\autoref{fig:single-surface}). There is no evidence for an interaction
between light treatment and surface (ANOVA,
\(F_{2,126} = 0.577, P = 0.563\)).
